---
title: "Explore Ornithology KE EMu database"
output: html_document
date: "2022-10-11"
editor_options: 
  chunk_output_type: console
---

Load libraries
```{r}
library(tidyverse)
library(lubridate)
library(ggplot2)
```



First install flatfile of EMu database received from Bill at 9/16/2022

```{r}
data <- read.csv("Birds_Collection.csv")

data <- data %>% 
  select(Catalog.No, Field.No, Sex, LAF.No, Age, Spec.Nat, Measurements, Gonads, Weight, Collector, Date.Coll, Family, Genus, Species, Subspecies, Continent, Country, State, County, Township, Nearest.Named.Place) %>% 
  mutate(lacm = Catalog.No,
         field = Field.No,
         sex = Sex,
         laf = LAF.No,
         age = Age, 
         specnat = Spec.Nat,
         measure = Measurements,
         gonads = Gonads,
         wt = as.numeric(Weight),
         coll = Collector,
         datecoll = Date.Coll,
         species = paste(Genus, Species, sep = " "),
         spp = Subspecies,
         genus = Genus,
         family = Family,
         locality = paste(Country, State, County, Township, Nearest.Named.Place, sep = " "),
         state = State,
         county = County,
         country = Country,
         continent = Continent
         ) %>% 
  select(lacm, field, sex, laf, age, specnat, measure, gonads, wt, coll, datecoll, family, species, genus, spp, locality, state, county, country, continent)



# transform date into an actual date category
data$date <- as.Date(data$datecoll, format="%d %B %Y")

min(data$date, na.rm=T)

data$year <- as.numeric(format(data$date, "%Y"))
data$month <- as.numeric(format(data$date, "%m"))

# remove odd dates
data2 <- data %>% filter(date > "1800-01-01")

dat <- data
```


Relative frequencies
```{r}
table(dat$family)
length(unique(dat$family)) #159 families
which.max(table(dat$family)) #Emberizidae

table(dat$specnat)

spectype <- dat %>% 
  group_by(specnat) %>% 
  summarise(n = n()) %>% 
  mutate(rel.freq = paste0(round(100* n/sum(n), 0), "%"))


fam <- dat %>% 
  group_by(family) %>% 
  summarise(n = n()) %>%
  mutate(rel.freq = paste0(round(100 * n/sum(n), 0), "%"))

# relative to species per family? 
length(unique(dat$species)) #5660

```


         AC    AL    AO    BM    FA    FB    FS    KB    MS    PS    SA    SK    SN    SO    SS    SW    TI    WO    WT 
   33    11  2872    15  1497    79  2063   126   989    62   417     7   214 15699   554 98160     3   794   133     2 

Frequencies by country
```{r}
# County
table(dat$country)

ctry <- dat %>% 
  group_by(country) %>% 
  summarise(n = n()) %>% 
  mutate(rel.freq = paste0(round(100* n/sum(n), 0), "%"))

# Continent
table(dat$continent)

cont <- dat %>% 
  group_by(continent) %>% 
  summarise(n = n()) %>% 
  mutate(rel.freq = paste0(round(100* n/sum(n), 0), "%"))

```


Frequenceis by time period
```{r}
dat2 <- dat %>% 
      mutate(Decade = as.numeric(year) - as.numeric(year) %% 10)

ggplot(dat2, aes(x=Decade)) +
  geom_histogram(binwidth = 3) + 
  theme_classic() +
  scale_x_continuous(name="year", limits = c(1870,2030), breaks = seq(1870,2030, by=10))

# by decade and continent
dat3 <- dat2 %>% 
  group_by(Decade, continent) %>% 
  summarize(n=n()) 

ggplot(dat3, aes(x=Decade, y=n, fill=continent)) +
  geom_col(position=position_dodge2(preserve="single", width=6), color="black", show.legend = T) +
  theme_classic() +
  theme(
    axis.text=element_text(size=14, color="black"),
    axis.title=element_text(size=14, color="black"),
    legend.text = element_text(size=14),
    legend.title = element_blank()) +
  scale_x_continuous(name="year", limits = c(1870,2020), breaks = seq(1870,2030, by=10))

# remove non-continents
dat4 <- dat3 %>% 
  filter(continent == "AFRICA" | continent == "EUROPE" | continent == "ASIA" | continent == "AUSTRALIA"
         | continent == "EUROPE" | continent == "NORTH AMERICA" | continent == "SOUTH AMERICA" | continent == "ANTARCTICA")

cbpal  <- c("#E69F00", "#56B4E9", "#009E73", 
                       "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

ggplot(dat4, aes(x=Decade, y=n, fill=continent)) +
  geom_col(position=position_dodge2(preserve="single", width=6), color="black", show.legend = T) +
  scale_fill_manual(name="continent", values=cbpal) +
  theme_classic() +
  theme(
    axis.text=element_text(size=14, color="black"),
    axis.title=element_text(size=14, color="black"),
    legend.text = element_text(size=14),
    legend.title = element_blank()) +
  scale_x_continuous(name="Decades", limits = c(1870,2020), breaks = seq(1870,2030, by=10)) +
  ylab("Count")

```



Plot some trends
```{r}
sp <- "Buteo jamaicensis"

example <- data5 %>% filter(species == sp)

table(example$year, example$specnat)
hist(example$date, breaks=20)

example2 <- example %>% 
  filter(specnat == "SS" | specnat == "SN")
min(example2$year)
max(example2$year)

# Histogram based on specimen nature
ggplot(example2, aes(x = year, fill = specnat, color = specnat)) +
  geom_histogram(breaks = seq(1880, 2020, by = 10), alpha = 0.5, position="identity") +
  scale_x_continuous(breaks = seq(1880, 2020, 10)) +
  theme_classic() +
  labs(fill = "Specimen type", color = "Specimen type", x = "Year", y = "Count",
       title = sp)

# Histogram based on subspecies
ggplot(example2, aes(x = year, fill = spp, color = spp)) +
  geom_histogram(breaks = seq(1880, 2020, by = 10), alpha = 0.5, position="dodge2") +
  scale_x_continuous(breaks = seq(1880, 2020, 10)) +
  theme_classic() +
  labs(fill = "Subspecies", color = "Subspecies", x = "Year", y = "Count",
       title = sp)
```



Add map of specimen count by county/state

devtools::install_github("UrbanInstitute/urbnmapr")
```{r}
library(urbnmapr)

# plot states
states_sf <- get_urbn_map("states", sf = TRUE)

states_sf %>%
  ggplot() +
  geom_sf(aes(), 
          fill = "grey", color = "#ffffff", size = 0.25) +
  geom_sf_text(data = get_urbn_labels(map = "states", sf = TRUE), 
                aes(label = state_abbv), size = 3)

# plot counties
counties_sf <- get_urbn_map("counties", sf = TRUE)

counties_sf %>% 
  ggplot(aes()) +
  geom_sf(fill = "grey", color = "#ffffff")


### create summary by county/state
## by state
statecount <- example2 %>% 
  count(state) %>% 
  mutate(state2 = substr(state, 1, 2)) #incorrect

spat_state <- left_join(get_urbn_map(map = "states", sf = TRUE),
                         statecount,
                         by = c("state_abbv" = "state2"))

# redo - correct 
statecount2 <- example %>% 
  count(abv) 

spat_state <- left_join(get_urbn_map(map = "states", sf = TRUE),
                         statecount2,
                         by = c("state_abbv" = "abv"))


# plot count by state
ggplot() +
  geom_sf(spat_state,
          mapping = aes(fill = n),
          color = "#ffffff", size = 0.25) +
  labs(fill = "Specimen count") +
  labs(title = paste(sp, "by state", sep = " "))


########################
## by county
countycount <- example2 %>% 
  count(state, county) %>% 
  mutate(state2 = substr(state, 1, 2)) %>% 
  mutate(statecounty = paste(state2, county, sep = "_"))

# need to capitalize and spell out County (County -> CO)
counties_sf$cty <- toupper(counties_sf$county_name)
counties_sf$cty2 <- gsub("COUNTY", "CO", counties_sf$cty)
counties_sf$statecounty <- paste(counties_sf$state_abbv, counties_sf$cty2, sep = "_")

# merge
spat_county <- left_join(counties_sf,
                         countycount,
                         by = "statecounty")

### works now 
# re-do within dataset
example2$cty <- str_to_title(example2$county) 
example2$cty2 <- gsub(" Co", " County", example2$cty)
countycount2 <- example2 %>% 
  count(cty2)

table(example2$cty2)
table(counties_sf$county_name)

counties_sf_ca <- get_urbn_map("counties", sf = TRUE) %>% 
  filter(state_abbv == "CA")

table(example2$cty2)
table(counties_sf_ca$county_name)


spat_county <- left_join(counties_sf_ca,
                         countycount2,
                         by = c("county_name"="cty2"))

###

# plot count by county
ggplot() +
  geom_sf(spat_county,
          mapping = aes(fill = n),
          color = "#ffffff", size = 0.25) +
  labs(fill = "Count by county") +
  labs(title = paste(sp, "by county", sep = " "))

```

Plot by county just WITHIN CALIFORNIA
```{r}
cali <- spat_county %>% filter(state_abbv == "CA")

ggplot() +
  geom_sf(cali,
          mapping = aes(fill = n),
          color = "#ffffff", size = 0.25) +
  labs(fill = "Count by county") +
  labs(title = paste(sp, "by CA county", sep = " "))
```



# Next steps 
Within LA County -> plot points / plot by cities / plot major 5 counties surrounding LA
Plot georeferences points ->note those that don't have georeference
table of total number of specimen types/sex/other fields.. 


```{r}
table(data$specnat)
```


Of the specimens we have, which one has the largest sampling area (by state #)?

```{r}
d <- data5 %>% 
  group_by(species) %>% 
  summarize(n = n(),
            count = n_distinct(abv))


```

